---
title: "Pattern Recognition With Tracker Data"
author: "Der-Hsuan Tsou"
date: "5/20/2021"
output: html_document
---

This is a Google Analytics Certificate Project. Data used in this project was retrieved from [Kaggle](https://www.kaggle.com/arashnic/fitbit).There are two main goals for this analysis:
**1.** Find the trend in the data   **2.**The usage of the trend.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Environment Set up
```{r}
# Set the D drive as working directory
setwd("D:/GoogleDataAnalytics/CaseStudy_BellaBeat/FitabaseData")
```
 
```{r}
#install.packages("tidyverse")
#install.packages('reshape2')
library(readr)
library(dplyr)
library(lubridate)
library(reshape2) #organize correlation heat map matrix
library(ggplot2)
```

## Import Data
```{r}
DailyAct <- read_csv('dailyActivity_merged.csv')      # Use
# DailyCalo <- read_csv('dailyCalories_merged.csv')     # No
# DailyInten <- read_csv('dailyIntensities_merged.csv') # No
# DailyStep <- read_csv('dailySteps_merged.csv')        # No

#HeartRateS <- read_csv('heartrate_seconds_merged.csv')# Keep
HourCalo <- read_csv('hourlyCalories_merged.csv')     # No
#HourInten <- read_csv('hourlyIntensities_merged.csv') # Keep
HourStep <- read_csv('hourlySteps_merged.csv')        # No

#MinMet <- read_csv('minuteMETsNarrow_merged.csv')     # Keep
#MinSleep <- read_csv('minuteSleep_merged.csv')        # Keep 
SleepDay <- read_csv('sleepDay_merged.csv')           # Use
WeightLog<- read_csv('weightLogInfo_merged.csv')      # Use
```


```{r}
summary(DailyAct)
```

*DailyCalo*, *DailyInten* and *DailyStep* are included in *DailyAct*, thus they will not be used in further analysis.

*HourStep*, *HourCalo* will be used for validation with *DailyAct*.
```{r}
#summary(DailyCalo)
#summary(DailyInten)
#summary(DailyStep)
#summary(HeartRateS)
summary(HourCalo)
#summary(HourInten)
#summary(MinMet) # Not sure what METs is
#summary(MinSleep) # Not sure what value and logid mean
summary(SleepDay) # *Analysis*: TotalSleepRecords and TotalMinutesAsleep TotalTimeInBed
summary(WeightLog)
```
Validate data with *DailyAct*

Sum of Daily Steps in *HourStep* is not exactly identical to *DailyAct* but they did not differentiate too much.
Sum of Daily Calories in *HourCalo* is not exactly identical to *DailyAct* but they did not differentiate too much.
```{r}
str(HourCalo)
HourCalo$ActivityHour <- as.Date(HourCalo$ActivityHour, format = "%m/%d/%Y")

HourCalo %>%
  group_by(Id, ActivityHour) %>%
  summarise(sum(Calories))
```

```{r}
str(HourStep)
HourStep$ActivityHour <- as.Date(HourStep$ActivityHour, format =  "%m/%d/%Y")

HourStep %>%
  group_by(Id, ActivityHour) %>%
  summarise(sum(StepTotal))
```
## Data Exploration
- change data type
- imputation
- plot to find pattern

##### DailyAct 

*TotalDistance* and *TrackerDistance* are basically identical.
```{r}
str(DailyAct)
DailyAct$Id <- as.factor(DailyAct$Id)
DailyAct$ActivityDate <- as.Date(DailyAct$ActivityDate, format = "%m/%d/%Y")
str(DailyAct)
head(DailyAct)
summary(DailyAct)
unique(DailyAct$Id) # 33 unique ID
```
build a *DailyAct* numeric variable correlation map
```{r}
DailyAct.num <- DailyAct %>%
  select(-c(Id,ActivityDate,TrackerDistance))
cormat <- round(cor(DailyAct.num),2)
head(cormat)
# get upper triangle of the correlation matrix
  get_upper_tri<-function(cormat){
    cormat[lower.tri(cormat)] <- NA
    return(cormat)
  }

# reorder the correlation matrix
reorder_cormat <- function(cormat){
# Use correlation between variables as distance
dd <- as.dist((1-cormat)/2)
hc <- hclust(dd)
cormat <-cormat[hc$order, hc$order]
}
# Reorder the correlation matrix
cormat <- reorder_cormat(cormat)
upper_tri <- get_upper_tri(cormat)
# Melt the correlation matrix
cormat_melt <- melt(upper_tri, na.rm = TRUE)
# Create a ggheatmap
ggheatmap <- ggplot(cormat_melt, aes(Var2, Var1, fill = value))+
 geom_tile(color = "white")+
 scale_fill_gradient2(low = "blue", high = "red", mid = "white", 
   midpoint = 0, limit = c(-1,1), space = "Lab", 
    name="Pearson\nCorrelation") +
  theme_minimal()+ # minimal theme
 theme(
   axis.title.x = element_blank(),
   axis.title.y = element_blank(),
   axis.text.x = element_text(angle = 45, vjust = 1,
                              size = 10.5, hjust = 1)) +
  coord_fixed() + 
  labs(title = 'DailyAct Numeric Variable Correlation')
print(ggheatmap)    
```

* Negative relationship happened between *SedentaryMinutes*, *LightActiveDistance* (-0.41) and *LightlyActiveMinutes*(-0.44)

* Positive relationship(0.95) happened between*ModeratelyActiveDistance* and *FairlyActiveMinutes* 

- slightly positive relationship(0.04) between *SedentaryActiveDistance* and *SedentaryMinutes*, which is normal

- positive relationship(0.89) shared between *LightActiveDistance* and *LightlyActiveMinutes* 

- positive relationship(0.83) shared between *VeryActiveDistance* and *VeryActiveMinutes*


##### SleepDay
```{r}
str(SleepDay)
SleepDay <- SleepDay %>%
  rename(SleepDatetime = SleepDay)
SleepDay$Id <- as.factor(SleepDay$Id)
# separate the date the time as the time are all the same, omit SleepDatetime column
SleepDay$SleepDate <- substr(SleepDay$SleepDatetime, 1, 9)
SleepDay$SleepDate <- as.Date(SleepDay$SleepDate, format = "%m/%d/%Y") #time only starts from 12AM
SleepDay$Weekday <- wday(SleepDay$SleepDate, label = TRUE, week_start = 1)
str(SleepDay)
head(SleepDay)
summary(SleepDay)
unique(SleepDay$Id) # 24 unique Id
```
Use bar chart to find the trend for *TotalMinutesAsleep*, *TotalTimeInBed* and *SleepDate*
```{r}
Slplt_Date <- SleepDay %>%
  select(TotalMinutesAsleep, TotalTimeInBed, SleepDate) %>%
  group_by(SleepDate) %>%
  summarise(TotalMinutesAsleep = mean(TotalMinutesAsleep), TotalTimeInBed = mean(TotalTimeInBed))
Slplt_Date02 <- melt(Slplt_Date, id.vars = 'SleepDate')
```

```{r, fig.height=5 , fig.width= 12}
plt <- ggplot(Slplt_Date02, aes(x = SleepDate, y = value, fill = variable)) +
  geom_bar(stat = 'identity', position = 'dodge') +
  scale_x_date(breaks = Slplt_Date02$SleepDate) +
  theme(axis.text.x = element_text(angle = 90, size = 10)) + 
  labs(y = 'Total Minutes', title = 'Time In Bed and Time Asleep Everyday')
plt
```

Limited info is found in this bar chart, thus below are zoomed in line chart between *TotalMinutesAsleep*, *TotalTimeInBed* and *SleepDate*. The line chart below can tell better difference between *TotalMinutesAsleep*, *TotalTimeInBed* and *SleepDate*.

```{r}
plt <- ggplot(data = Slplt_Date, aes(x = SleepDate)) + 
  geom_line(aes(y = TotalMinutesAsleep), color = 'orange') + 
  geom_line(aes(y = TotalTimeInBed), color = 'dark green') +
  labs(y = 'Total Minutes', title = "Time In Bed and Time Asleep Everyday") +
  scale_x_date(breaks = Slplt_Date02$SleepDate) +
  theme(axis.text.x = element_text(angle = 90, size = 8)) +
  geom_hline(yintercept = 397.8, color = 'pink') + 
  geom_hline(yintercept = 478.2, color = 'pink') +
  annotate("text", x= as.Date('2016-04-16'), y=390, label = "Sat", 
           color ='red', size=3) +
  annotate("text", x= as.Date('2016-04-21'), y=374.5, label = "Thu", 
           color ='black', size=3) +
  annotate("text", x= as.Date('2016-04-26'), y=368, label = "Tue", 
           color ='black', size=3) + 
  annotate("text", x= as.Date('2016-04-29'), y=384.5, label = "Fri", 
           color ='black', size=3) +
  annotate("text", x= as.Date('2016-05-03'), y=395.5, label = "Tue", 
           color ='black', size=3) +
  annotate("text", x= as.Date('2016-05-05'), y=368.5, label = "Thu", 
           color ='black', size=3) +
  annotate("text", x= as.Date('2016-04-12'), y=483, label = "Tue", 
           color ='black', size=3) +
  annotate("text", x= as.Date('2016-04-14'), y=483.5, label = "Thu", 
           color ='black', size=3) +
  annotate("text", x= as.Date('2016-04-17'), y=511, label = "Sat", 
           color ='black', size=3) +
  annotate("text", x= as.Date('2016-04-24'), y=538, label = "Sun", 
           color ='black', size=3) +
  annotate("text", x= as.Date('2016-04-24'), y=489, label = "Sun", 
           color ='red', size=3) +
  annotate("text", x= as.Date('2016-04-30'), y=498, label = "Sat", 
           color ='red', size=3) +
  annotate("text", x= as.Date('2016-05-01'), y=488, label = "Sun", 
           color ='red', size=3) +
  annotate("text", x= as.Date('2016-05-02'), y=495, label = "Mon", 
           color ='red', size=3) +
  annotate("text", x= as.Date('2016-05-08'), y=488, label = "Sun", 
           color ='black', size=3)
plt

```

The pink lines shown in the line chart above are the 25% quantile of *TotalTimeInBed* and 75% quantile of *TotalTimeAsleep*.

- Among those test people, they spent the most time in bed and also the most time asleep on Sunday, April, 24th, 2020

- Three days in a row on the weekend from April, 30th to May, 2nd, 2020, people spent time in bed a lot, however, asleep less. I am assuming it is a holiday weekend but further analysis are required due to lack of the location of the people

- I assumed less asleep time will happen on weekdays, however, on Sat, April, 16th, 2020, people had *TotalAsleepTime* less than the 25% average

The chart for days in the week were plotted below for deeper analysis.

```{r}
Slplt_Weekday <- SleepDay %>%
  select(TotalMinutesAsleep, TotalTimeInBed, Weekday) %>%
  group_by(Weekday) %>%
  summarise(TotalMinutesAsleep = mean(TotalMinutesAsleep), TotalTimeInBed = mean(TotalTimeInBed))
Slplt_Weekday02 <- melt(Slplt_Weekday, id.vars = 'Weekday')
```

```{r}
plt <- ggplot(data = Slplt_Weekday02, aes(x = Weekday, y = value, fill = variable)) + 
  geom_bar(stat = 'identity', position = 'dodge') +
  labs(x = "Day Of Week", y = 'Total Minutes', title = "Time In Bed and Time Asleep In Day of Week")
plt
```

I thought people will spend more time in bed and sleep more on the weekends, however, our data show a different story. They tend to sleep the most on Sunday and Wednesday. This raises my curiosity in what kind of occupation the 24 people are.


##### WeightLog
```{r}
str(WeightLog)
WeightLog$Id <- as.factor(WeightLog$Id)
WeightLog$RecordDate <- substr(WeightLog$Date, 1, 9)
WeightLog$RecordDate <- as.Date(WeightLog$RecordDate, format = '%m/%d/%Y')
WeightLog$Date <- mdy_hms(WeightLog$Date, tz=Sys.timezone())
str(WeightLog)
summary(WeightLog) #65/67 NA in Fat so will not be used for further analysis
WeightLog <- WeightLog %>%
  select(-Fat)
unique(WeightLog$Id) # 8 unique Id
```

```{r}
df <- WeightLog %>%
  select(RecordDate, WeightPounds, BMI) %>%
  group_by(RecordDate) %>%
  summarise(AVGWeight = mean(WeightPounds), AVGBMI = mean(BMI))
```

```{r}
plt <- ggplot(data = df, aes(x= RecordDate)) +
  geom_line(aes(y = AVGWeight), color = 'purple') +
  geom_line(aes(y = AVGBMI), color = 'blue') +
  labs(y = 'BMI & Pounds') +
  scale_x_date(breaks = df$RecordDate) +
  theme(axis.text.x = element_text(angle = 90, size = 8)) +
  geom_point(aes(x = as.Date('2016-04-13'), y = 206.1322), color = 'red') +
  geom_point(aes(x = as.Date('2016-04-13'), y = 32.39667), color = 'red') +
  geom_point(aes(x = as.Date('2016-04-26'), y = 187.6134), color = 'red') +
  labs(title = 'Average Weight(LB) and BMI Trend')
plt
```

Based on the line chart above, there are three data points marked in red on April, 13th and April, 26th, which are abnormal and extremely high. Further analysis will be conducted.


Below, I will drill deeper to see the potential reason for both *WeightPounds* and *BMI* being high on April, 13th and *WeightPounds* High but *BMI* normal on April, 26th.
```{r}
Weight_df <- WeightLog %>%
  select(Id, RecordDate, WeightPounds, BMI) %>%
  group_by(Id) %>%
  count(Id) %>%
  arrange(Id)
Weight_df
df <- WeightLog %>%
  select(Id, RecordDate, WeightPounds, BMI) %>%
  filter(RecordDate == '2016-04-13'| RecordDate == '2016-04-26')
df
```
According to two of the zoomed in data, which the first one grouped and counted by the *Id* ; the second one filtered on only the two days we cared about(4/13 & 4/26), I am guessing two facts:

1. *Id* 1927972279 has obesity and *Id* 1927972279 has only one record, which led to both *WeightPounds* and *BMI* raise on April 13th

2. *Id* 8877689391 is tall, which caused only the *WeightPounds* raised. Besides, April 26th has only one record, which is *Id* 8877689391. That explained although *Id* 8877689391 has total 24 records in the *WeightLog* data, only April 26th was affected.

- The above assumptions are based on the information I have, further validation would need more data such as height.
```{r}

```









## Reference
FitBit Fitness Tracker Data. Kaggle. [https://www.kaggle.com/arashnic/fitbit](https://www.kaggle.com/arashnic/fitbit).

ggplot2 : Quick correlation matrix heatmap - R software and data visualization. STHDA.
http://www.sthda.com/english/wiki/ggplot2-quick-correlation-matrix-heatmap-r-software-and-data-visualization
